---
title: "SPX Report"
format:
  html:
    toc: true
    embed-resources: true
engine: knitr
---

```{r Load R libraries, echo = FALSE}
library(reticulate)
library(tidyverse)
library(pomp)

setwd(file.path(here::here(), "spx", "performance"))
sessionInfo()

knitr::opts_chunk$set(echo = FALSE)
```

```{r}
print(Sys.time())
```

```{python Import python modules}
import os
import pickle
import numpy as np
import pandas as pd
import pypomp as pp
import jax
```

```{python Load data in python}
PKL_PATH = "performance/gpu_results/spx_results_rl4.pkl"
pkl_in = pickle.load(open(PKL_PATH, "rb"))
```

```{python Make basic frame}
LL_frame = pkl_in.results()
LL_frame = (
    pd.DataFrame(
        {
            "LL": LL_frame["logLik"],
            "sd": LL_frame["se"],
        }
    )
    .sort_values(by="LL", ascending=False)
    .reset_index(drop=True)
)

traces = pkl_in.traces()
```

# Python results

## Process results

```{r Make traces frame in R}
traces <- py$traces
LL_frame <- py$LL_frame
best_rep <- LL_frame$index[[1]]
traces <- traces |>
    rename(rep = replicate, iter = iteration) |>
    pivot_longer(
        cols = c(-rep, -iter, -logLik, -method),
        names_to = "quantity",
        values_to = "param_value"
    )
```

## Plot traces

```{r Set best_rep}
# Set best_rep to the rep with the largest logLik on the last iteration
last_iter <- max(traces$iter)
best_rep <- traces |>
    filter(iter == last_iter, method == "pfilter") |>
    group_by(rep) |>
    summarise(final_logLik = logLik[1]) |>
    arrange(desc(final_logLik)) |>
    slice(1) |>
    pull(rep)
```

```{r Plot traces}
ggplot(
    filter(traces, quantity != "logLik"),
    aes(x = iter, y = param_value, group = rep, color = as.factor(rep))
) +
    geom_line(show.legend = FALSE) +
    facet_wrap(vars(quantity), scales = "free")
```

```{r Plot LL traces, eval = FALSE}
ggplot(
    filter(traces, quantity == "logLik"),
    aes(x = iter, y = param_value, group = rep, color = as.factor(rep))
) +
    geom_line() +
    facet_wrap(vars(quantity), scales = "free") +
    coord_cartesian(ylim = c(11000, 11850))
```

#  Compare python and R traces

```{r Load R estimates}
load(file.path(
    "performance",
    "R_results",
    "search360_hidden",
    "1d_global_search360.rda"
))
LL_df_R <- as.data.frame(L.box)
traces_long <- lapply(seq_along(if.box), function(x) {
    tr <- traces(if.box[[x]])
    tr_long <- tr |>
        as.data.frame() |>
        mutate(iter = 1:nrow(tr), rep = x) |>
        pivot_longer(
            names_to = "quantity",
            values_to = "param_value",
            cols = c(-iter, -rep)
        )
}) |>
    bind_rows() |>
    mutate(language = "R") |>
    mutate(
        value_T = case_when(
            quantity == "logLik" ~ param_value,
            quantity == "mu" ~ log(param_value),
            quantity == "kappa" ~ log(param_value),
            quantity == "theta" ~ log(param_value),
            quantity == "xi" ~ log(param_value),
            quantity == "rho" ~ log((1 + param_value) / (1 - param_value)),
            quantity == "V_0" ~ log(param_value)
        )
    )

traces_pyr <- bind_rows(
    traces |>
        mutate(language = "python") |>
        mutate(
            value_T = case_when(
                quantity == "logLik" ~ param_value,
                quantity == "mu" ~ log(param_value),
                quantity == "kappa" ~ log(param_value),
                quantity == "theta" ~ log(param_value),
                quantity == "xi" ~ log(param_value),
                quantity == "rho" ~ log((1 + param_value) / (1 - param_value)),
                quantity == "V_0" ~ log(param_value),
                TRUE ~ param_value
            )
        ),
    traces_long
)
```

```{r Compare python and R traces}
traces_pyr |>
    filter(quantity != "logLik") |>
    filter(value_T >= -12, value_T <= 12) |>
    ggplot(
        mapping = aes(
            x = iter,
            y = value_T,
            group = interaction(rep, language),
            color = as.factor(language)
        )
    ) +
    geom_line(alpha = 0.2) +
    facet_wrap(vars(quantity), scales = "free") +
    ggtitle("Comparison of Python and R Parameter Traces")
```

```{r Compare python and R traces quantiles}
traces_pyr |>
    filter(quantity != "logLik") |>
    filter(value_T >= -12, value_T <= 12) |>
    group_by(iter, quantity, language) |>
    summarise(
        ymin = quantile(value_T, 0.1, na.rm = TRUE),
        ymax = quantile(value_T, 0.9, na.rm = TRUE),
        .groups = "drop"
    ) |>
    ggplot(
        mapping = aes(x = iter, ymin = ymin, ymax = ymax, fill = language)
    ) +
    geom_ribbon(alpha = 0.5) +
    facet_wrap(vars(quantity), scales = "free") +
    ggtitle("Comparison of Python and R Parameter Trace Quantiles")
```

# Compare python and R estimates

```{r Compare python and R estimates}
traces_pyr |>
    group_by(language, quantity) |>
    filter(iter == max(iter)) |>
    summarise(
        p10 = quantile(value_T, 0.1, na.rm = TRUE),
        p90 = quantile(value_T, 0.9, na.rm = TRUE),
        .groups = "drop"
    ) |>
    right_join(
        traces_pyr |>
            group_by(language) |>
            filter(iter == max(iter)),
        by = c("language", "quantity")
    ) |>
    filter(value_T >= -10, value_T <= 10) |>
    ggplot(mapping = aes(x = value_T, fill = language)) +
    geom_density(alpha = 0.5) +
    facet_wrap(vars(quantity), scales = "free") +
    labs(
        title = "Comparison of Final Iteration Values for Each Parameter (Zoomed)",
        x = "Parameter",
        y = "Transformed Value",
        fill = "Language"
    )
```

```{r Compare python and R estimates (boxplot)}
traces_pyr |>
    filter(quantity != "logLik", quantity != "loglik") |>
    group_by(language, quantity) |>
    filter(iter == min(iter)) |>
    ggplot(mapping = aes(x = value_T, fill = language)) +
    geom_density(alpha = 0.5) +
    facet_wrap(vars(quantity), scales = "free") +
    labs(
        title = "Comparison of Starting Parameter Values for Each Parameter (Zoomed)",
        x = "Parameter",
        y = "Transformed Value",
        fill = "Language"
    )
```

## LL comparison

### Python LL summary:
```{r}
print(summary(py$LL_frame$LL))
head(arrange(py$LL_frame, desc(LL)), 10)
```

### R LL summary:
```{r}
print(summary(LL_df_R$est))
head(arrange(LL_df_R, desc(est)), 10)
```

### Density plot comparison

```{r}
data.frame(
    python = py$LL_frame$LL,
    R = LL_df_R$est
) |>
    filter(python > 11000, R > 11000) |>
    pivot_longer(
        names_to = "language",
        values_to = "LL",
        cols = c(python, R)
    ) |>
    ggplot(aes(x = LL, fill = language)) +
    geom_density(alpha = 0.5) +
    labs(
        title = "Comparison of Python and R LL Estimates (Zoomed)",
        x = "LL",
        fill = "Language"
    )
```

# Compare python and R LL estimates at Weizhe's estimates

```{python Load rep3600 eval results}
EVAL_PKL_PATH = os.path.join("pfilter_check", "py_results", "spx_results_eval.pkl")
pkl_eval = pickle.load(open(EVAL_PKL_PATH, "rb"))
LL_rep3600 = pkl_eval.results_history[0].logLiks[0].values.tolist()
```

```{r Load rep3600 R eval and plot combined density}
load(file.path("pfilter_check", "R_results", "spx_results_eval.rda"))

# Extract numeric log-likelihoods from L.box regardless of shape
LL_rep3600_R <- as.numeric(L.box)

data.frame(
    LL = c(unlist(py$LL_rep3600), LL_rep3600_R),
    source = c(
        rep("python", length(py$LL_rep3600)),
        rep("R", length(LL_rep3600_R))
    )
) |>
    ggplot(aes(x = LL, fill = source)) +
    geom_density(alpha = 0.5) +
    labs(
        title = "rep3600 Log-Likelihood Density: Python vs R",
        x = "Log-Likelihood",
        fill = "Source"
    )
```

# Weizhe's estimates

This table compares the parameter estimates reported by Weizhe in his thesis to our best python parameter estimates.

```{r Weizhe estimates}
rho_transform <- function(x) {
    out <- log((1 + x) / (1 - x))
    return(out)
}

# Weizhe values
data <- list(
    Parameter = c("mu", "kappa", "theta", "xi", "rho", "V_0"),
    Value = c(
        log(3.68e-4),
        log(3.14e-2),
        log(1.12e-4),
        log(2.27e-3),
        rho_transform(-7.38e-1),
        log(7.66e-3^2)
    )
)
df <- as.data.frame(data)

# Compare Weizhe's estimates to the best python fit
traces |>
    filter(iter == last_iter, rep == best_rep, method == "pfilter") |>
    select(quantity, param_value) |>
    mutate(
        param_value_T = case_when(
            quantity == "logLik" ~ param_value,
            quantity == "mu" ~ log(param_value),
            quantity == "kappa" ~ log(param_value),
            quantity == "theta" ~ log(param_value),
            quantity == "xi" ~ log(param_value),
            quantity == "rho" ~ log((1 + param_value) / (1 - param_value)),
            quantity == "V_0" ~ log(param_value),
            TRUE ~ param_value
        )
    ) |>
    left_join(df, by = c("quantity" = "Parameter")) |>
    mutate(difference = param_value_T - Value) |>
    rename(Weizhe_value = Value, python_value = param_value_T) |>
    print()
```
