---
title: "SPX Report"
format:
  html:
    toc: true
    embed-resources: true
engine: knitr
---

```{r Load R libraries, echo = FALSE}
library(reticulate)
library(tidyverse)
library(pomp)
setwd(paste0(here::here(), "/spx"))
sessionInfo()

knitr::opts_chunk$set(echo = FALSE)
```

```{python Import python modules}
import os
import pickle
import numpy as np
import pandas as pd
import pypomp as pp
import jax
```

```{python Load data in python}
PKL_PATH = "gpu/v2/spx_results.pkl"
pkl_in = pickle.load(open(PKL_PATH, "rb"))
```

```{python Make basic frame}
LL_frame = pkl_in.results()
LL_frame = (
    pd.DataFrame(
        {
            "LL": LL_frame["logLik"],
            "sd": LL_frame["se"],
        }
    )
    .sort_values(by="LL", ascending=False)
    .reset_index(drop=True)
)

traces = pkl_in.traces()
```

# Python results

## Process results

```{r Make traces frame in R}
traces <- py$traces
LL_frame <- py$LL_frame
best_rep <- LL_frame$index[[1]]
traces <- traces |>
    rename(rep = replication, iter = iteration, logLik = loglik) |>
    pivot_longer(
        cols = c(-rep, -iter, -logLik, -method),
        names_to = "quantity",
        values_to = "param_value"
    )
```

## Plot traces

```{r Set best_rep}
# Set best_rep to the rep with the largest logLik on the last iteration
last_iter <- max(traces$iter)
best_rep <- traces |>
    filter(iter == last_iter, method == "pfilter") |>
    group_by(rep) |>
    summarise(final_logLik = logLik[1]) |>
    arrange(desc(final_logLik)) |>
    slice(1) |>
    pull(rep)
```

```{r Plot traces}
ggplot(
    filter(traces, quantity != "logLik"),
    aes(x = iter, y = param_value, group = rep, color = as.factor(rep))
) +
    geom_line(show.legend = FALSE) +
    facet_wrap(vars(quantity), scales = "free")
```

```{r Plot LL traces, eval = FALSE}
ggplot(
    filter(traces, quantity == "logLik"),
    aes(x = iter, y = param_value, group = rep, color = as.factor(rep))
) +
    geom_line() +
    facet_wrap(vars(quantity), scales = "free") +
    coord_cartesian(ylim = c(11000, 11850))
```

#  Compare python and R traces

```{r Load R estimates}
load("R/1d_global_search.rda")
LL_df_R <- as.data.frame(L.box)
traces_long <- lapply(seq_along(if.box), function(x) {
    tr <- traces(if.box[[x]])
    tr_long <- tr |>
        as.data.frame() |>
        mutate(iter = 1:nrow(tr), rep = x) |>
        pivot_longer(
            names_to = "quantity", values_to = "param_value", cols = c(-iter, -rep)
        )
}) |>
    bind_rows() |>
    mutate(language = "R") |>
    mutate(value_T = case_when(
        quantity == "logLik" ~ param_value,
        quantity == "mu" ~ log(param_value),
        quantity == "kappa" ~ log(param_value),
        quantity == "theta" ~ log(param_value),
        quantity == "xi" ~ log(param_value),
        quantity == "rho" ~ log((1 + param_value) / (1 - param_value)),
        quantity == "V_0" ~ log(param_value)
    ))

traces_pyr <- bind_rows(
    traces |> mutate(language = "python", value_T = param_value),
    traces_long
)
```

```{r Compare python and R traces}
traces_pyr |>
    filter(quantity != "logLik") |>
    filter(value_T >= -12, value_T <= 12) |>
    ggplot(
        mapping = aes(
            x = iter, y = value_T, group = interaction(rep, language),
            color = as.factor(language)
        )
    ) +
    geom_line(alpha = 0.2) +
    facet_wrap(vars(quantity), scales = "free") +
    ggtitle("Comparison of Python and R Parameter Traces")
```

```{r Compare python and R traces quantiles}
traces_pyr |>
    filter(quantity != "logLik") |>
    filter(value_T >= -12, value_T <= 12) |>
    group_by(iter, quantity, language) |>
    summarise(
        ymin = quantile(value_T, 0.1, na.rm = TRUE),
        ymax = quantile(value_T, 0.9, na.rm = TRUE),
        .groups = "drop"
    ) |>
    ggplot(
        mapping = aes(x = iter, ymin = ymin, ymax = ymax, fill = language)
    ) +
    geom_ribbon(alpha = 0.5) +
    facet_wrap(vars(quantity), scales = "free") +
    ggtitle("Comparison of Python and R Parameter Trace Quantiles")
```

## Compare python and R estimates

```{r Compare python and R estimates}
traces_pyr |>
    group_by(language, quantity) |>
    filter(iter == max(iter)) |>
    summarise(
        p10 = quantile(value_T, 0.1, na.rm = TRUE),
        p90 = quantile(value_T, 0.9, na.rm = TRUE),
        .groups = "drop"
    ) |>
    right_join(
        traces_pyr |>
            group_by(language) |>
            filter(iter == max(iter)),
        by = c("language", "quantity")
    ) |>
    filter(value_T >= -10, value_T <= 10) |>
    ggplot(mapping = aes(x = language, y = value_T, color = language), alpha = 0.2) +
    geom_jitter() +
    facet_wrap(vars(quantity), scales = "free") +
    labs(
        title = "Comparison of Final Iteration Values for Each Parameter (Zoomed)",
        x = "Parameter",
        y = "Transformed Value",
        fill = "Language"
    )
```

```{r Compare python and R estimates (boxplot)}
traces_pyr |>
    filter(quantity != "logLik", quantity != "loglik") |>
    group_by(language, quantity) |>
    filter(iter == min(iter)) |>
    ggplot(mapping = aes(x = language, y = value_T, color = language), alpha = 0.2) +
    geom_boxplot() +
    facet_wrap(vars(quantity), scales = "free") +
    labs(
        title = "Comparison of Starting Parameter Values for Each Parameter (Zoomed)",
        x = "Parameter",
        y = "Transformed Value",
    )
```

## LL comparison

### Python LL summary:
```{r}
print(summary(py$LL_frame$LL))
head(arrange(py$LL_frame, desc(LL)), 10)
```

### R LL summary:
```{r}
print(summary(LL_df_R$est))
head(arrange(LL_df_R, desc(est)), 10)
```

### Density plot comparison

```{r}
data.frame(
    python = py$LL_frame$LL,
    R = LL_df_R$est
) |>
    filter(python > 11000, R > 11000) |>
    pivot_longer(names_to = "language", values_to = "LL", cols = c(python, R)) |>
    ggplot(aes(x = LL, fill = language)) +
    geom_density(alpha = 0.5) +
    labs(
        title = "Comparison of Python and R LL Estimates (Zoomed)",
        x = "LL",
        fill = "Language"
    )
```

# Weizhe's estimates

This table compares the parameter estimates reported by Weizhe in his thesis to our best python parameter estimates.

```{r Weizhe estimates}
rho_transform <- function(x) {
    out <- log((1 + x) / (1 - x))
    return(out)
}

# Weizhe values
data <- list(
    Parameter = c("mu", "kappa", "theta", "xi", "rho", "V_0"),
    Value = c(
        log(3.68e-4),
        log(3.14e-2),
        log(1.12e-4),
        log(2.27e-3),
        rho_transform(-7.38e-1),
        log(7.66e-3^2)
    )
)
df <- as.data.frame(data)

# Compare Weizhe's estimates to the best python fit
traces |>
    filter(iter == last_iter, rep == best_rep, method == "pfilter") |>
    select(quantity, param_value) |>
    left_join(df, by = c("quantity" = "Parameter")) |>
    mutate(difference = param_value - Value) |>
    rename(Weizhe_value = Value, python_value = param_value) |>
    print()
```

Here we check that, when running the particle filter at the values reported by Weizhe in his thesis, we obtain a LL estimate similar to the reported value and similar to what we obtained from running the particle filter in R.

```{python} 
def rho_transform(x):
    return np.log((1 + x) / (1 - x))


key = jax.random.key(1)
spx_obj = pp.spx()

key, subkey = jax.random.split(key)
spx_obj.pfilter(
    J=1000,
    key=subkey,
    theta={
        "mu": np.log(3.68e-4),
        "kappa": np.log(3.14e-2),
        "theta": np.log(1.12e-4),
        "xi": np.log(2.27e-3),
        "rho": rho_transform(-7.38e-1),
        "V_0": np.log(7.66e-3**2),
    },
    reps=72,
)

spx_obj.results()
```

Weizhe obtained an LL estimate of 11847.12 from profiling according to his thesis

Evaluating the Weizhe model with the Weizhe parameter estimates under Honor_Thesis/output/1d_global/weizhe_eval yielded LL 11849.65, sd 1.1217372.

